Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-02-22 draft.tex
.

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
%\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage{geometry}
\geometry{
a4paper,
left=2.0in,
right=1.0in}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}

@@

\subsubsection{Mean outgoing Weights}\label{Section_Mean_outgoing_Weights}
Since we only implemented synaptic normalization for the sum of \emph{ingoing} weights, the total or average strength of \emph{outgoing} connections may vary from cell to cell. In a sense, the strength of outgoing connections per neuron can be regarded as a measure of how "influential" a neuron's activity is with respect to other neurons in the network. Effenberger et al. have shown in computational studies that these "driver neurons" form highly active and interconnected subnetworks \cite{Effenberger_2015}, an observation that is backed up by experimental studies \cite{Yassin_Subnetworks_2010,Eckmann_Leader_Neurons_2008}. Figure \ref{Out_Weight_Mean_Quantile} shows that diffusive homeostasis allows for the emergence of a small group of highly influential neurons, which reproduces the findings in \cite{Effenberger_2015}. A comparison of statistics generated with shuffled versions of the weight matrices illustrates that above-chance exceptionally strong outgoing weights were indeed present. We noted further that the difference between shuffled matrices corresponding to non-diffusive and diffusive homeostasis is in line with our observation from Figure \ref{Weight_Dist}, which indicated a broader distribution of weights for diffusive homeostasis.
\begin{figure}
%%% plot_sum_outgoing_weights_dist.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/outgoing_weights_distribution.png}
\caption{Mean of outgoing weights of excitatory neurons in order sorted by magnitude. Dashed lines are means obtained by randomly shuffling synaptic weights. Data was taken from 25 weight matrices per line, representing weights within $\mathrm{t=1000-1500\,s}$.}
\label{Out_Weight_Mean_Quantile}
\end{figure}

Effenberger et al. identified the presence of inhibitory STDP as a necessary mechanism for the development of this feature, in contrast to our model. On a deeper explanatory level, the authors argue that inhibitory STDP is indirectly involved: It weakens inhibitory currents for some neurons, which cause these neurons to have more above-average firing rates. In turn, synapses between highly active presynaptic cells and postsynaptic neurons with lower activity are known to be subject to long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}. With these causal relations in mind, we can argue that diffusive homeostasis effectively embodies the same functional role as inhibitory STDP in \cite{Effenberger_2015} by allowing for the presence of excitatory cells with above-average activity. We tested this relationship by plotting the mean outgoing weights against the average firing rate, see Figure \ref{Out_Weight_vs_F}. A strong heterogeneity of firing rates allows for the development of few strong synapses, while the distribution of mean weights resulting from a narrow distribution is limited to a smaller range. We also tested the case of instant diffusion. Though firing rates span across a similar range, mean weights do not reach as high values as in the case of diffusion at a finite rate.
\begin{figure}
%%% out_weight_vs_f.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/mean_out_weights_vs_rate.png}
\caption{Mean of outgoing excitatory weights at $\mathrm{t=1500\,s}$ plotted against mean presynaptic firing rate (averaged over $\mathrm{t=700-1000\,s}$) in regular (A) and logarithmic space (B). Each plot corresponds to a presynaptic excitatory cell.}
\label{Out_Weight_vs_F}
\end{figure}
Furthermore, it is interesting to note that we found a roughly linear relationship on a logarithmic weight scale. Figure \ref{Out_Weight_vs_F} only depicts data from simulations including a distance dependent connection profile, since removing this topology did not result in noticeable differences.
\newpage
\section{Further Analysis of the Firing Rate Distribution}
The results of section \ref{Fir_Dist_Section} were positive on a preliminary level regarding our goal of achieving a broad and skewed firing rate distribution. Following this, we tried to further gain insight into the actual shape of the resulting distribution. 
\subsection{Review of a Dynamic Mean-Field Model}\label{Mean_Field_Review_Section}
Sweeney et al. attempted to predict the shape of the firing rate distribution by means of a simplified mean-field approximation of diffusion \cite{Sweeney_Paper}. The main simplification within homeostatic feedback was the replacement of Equation \ref{Theta_dyn} by
\begin{equation}
\dot{V_{t,i}}(t) = \frac{1}{\tau_{HIP}} \left( (1-\alpha)\frac{\phi_i-\phi_0}{\phi_i} +\alpha \frac{\langle \phi \rangle-\phi_0}{\langle \phi \rangle} \right)\;. 
\label{diff_hom_simpl_sweeney}
\end{equation}
$\mathrm{\alpha \: \epsilon \: [0,1]}$ thereby acted as a parameter that determines the mixture between single-neuron homeostasis ($\mathrm{\alpha=0}$) and the limit of quasi instantaneous spreading of the diffusive signal across the population ($\mathrm{\alpha=1}$). $\mathrm{\phi_i}$ and $\mathrm{\mathrm{\langle \phi \rangle}}$ represent individual firing rates and their overall mean.

@@

Generally speaking, no matter how the actual shape of the numeric solution in the equilibrium at a constant production rate looks like, it must be of the form
\begin{align}
NO_i(\mathbf{x}) &= r_i \cdot \psi (d(\mathbf{x}_{neur,i},\mathbf{x})) \label{general_diff_interaction} \\
d(\mathbf{x},\mathbf{y}) &\equiv |\mathbf{x}-\mathbf{y}| \; . \label{eucl_dist}
\end{align}
The full solution is then
\begin{equation}
NO(\mathbf{x}) = \sum_i NO_i(\mathbf{x})\;.
\label{full_sol_diff_equil}
\end{equation}
By defining
\begin{equation}
\psi_{ij} \equiv \psi_{ij} \equiv \psi (d(\mathbf{x}_{neur,i},\mathbf{x}_{neur,i}))
\label{interact_matrix_elements}
\end{equation}
we could express the condition \eqref{NO_equil_cond} as
\begin{equation}
\sum_j \psi_{ij}\cdot r_j = NO_0
\label{NO_equil_cond_interact_matrix}
\end{equation}
or, as an operator
\begin{align}
\hat{\psi}\mathbf{r} &= NO_0 \mathbf{n} \label{NO_equil_cond_interact_matrix_operator} \\
\mathbf{n}&\equiv (1,1,...,1) \; .
\end{align}
The problem of finding the steady-state solution of the homeostatic constraint thus reduced to inverting $\hat{\psi}$:
\begin{equation}
\mathbf{r} = NO_0 \hat{\psi}^{-1} \mathbf{n}
\label{NO_euqil_cond_interact_matrix_operator_solve}
\end{equation}
Still, we had to find a modified, non-diverging version of $\mathrm{\psi (d(\mathbf{x}_{neur,i},\mathbf{x}))}$ to acquire any prediction from this model. It had to retain the shape given by \eqref{solution_diff_equil_bessel} for larger distances but approach the correct numeric error value at the origin, determined by the spacing of the numeric grid. We solved this problem by the following expression:
\begin{equation}
\psi_{approx.} \equiv \frac{1}{\left(\frac{1}{\psi_0^\varepsilon} + \frac{1}{\psi_{point}^\varepsilon}\right)^{\frac{1}{\varepsilon}}}
\label{Numeric_Solution_Expression_Trick}
\end{equation}
where $\mathrm{\varepsilon}$ determines the "smoothness" of transition between $\mathrm{\psi}$ and the cutoff value $\mathrm{\psi_0}$. We chose $\mathrm{\varepsilon=10}$ for all further calculations. We took the simple approach of interpreting this value as a mean of the analytic solution across the area covered by the corresponding grid cell to find an expression for $\mathrm{\psi_0}$. As an additional simplification, we substituted the necessary integration over the square grid cell by a circular area of equal size around the source. This calculation yielded
\begin{equation}
\psi_0 = \gamma \frac{1-h\sqrt{\frac{\lambda}{\pi D}} K_1\left( h\sqrt{\frac{\lambda}{\pi D}}\right) }{h^2 \lambda}
\label{Numeric_Grid_Bessel_Approx}
\end{equation}
where $\mathrm{h}$ is the spatial resolution of the grid cells. Figure \ref{Psi_Approximation_Compare} shows a comparison between the numerically calculated solution and the expression given by \ref{Numeric_Solution_Expression_Trick}. The approximation fitted very well for nonzero values, as expected. The value at the critical point in the origin was slightly underestimated by \eqref{Numeric_Grid_Bessel_Approx}, but nonetheless fitted well into the overall shape.
\begin{figure}
%% compare_num_diff_steady_vs_analytic.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/Psi_Interact_Approximation.png}
\end{center}
\caption{Numerical steady-state solution of Equation \eqref{simple_NO_dyn_with_diff_equil_helmholtz} (blue, $\mathrm{h=10\;\mu m}$) and its approximation (dashed) given by \eqref{Numeric_Solution_Expression_Trick}. The numerical simulation was carried out on a 2d-grid and the curve represents a cut through the origin in x-direction.}
\label{Psi_Approximation_Compare}
\end{figure}
We took this approximation as a basis for further calculations of the interaction-matrix $\mathrm{\hat{\psi}}$.

By simply calculating all matrix elements of $\mathrm{\hat{\psi}}$ by means of Equation \eqref{interact_matrix_elements}, one would neglect the finite boundaries of the system, which would cause neurons close to the edge to "bleed" into empty space. This in turn would cause the solution of \eqref{NO_equil_cond_interact_matrix_operator} to contain an over-representation of high firing rates since close-to-the-edge neurons would need to compensate for their lack of neighbors. Thus, we had to account for the boundary conditions used in the network simulation. We simulated the network mostly with Neumann boundary conditions as well as periodic boundaries, as described in section \ref{methods}. Both types can be modeled by extending the neurons' population through spatially shifted and mirrored versions of the base population (see Figure \ref{Bound_Cond_Patches}):
\begin{itemize}
\item Periodic boundary conditions are induced by copies of the neurons' positions shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$.
\item Zero flux through the boundaries can be achieved by copied positions being shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$ \textit{and} mirrored in $x$($y$)-direction if $n_x$($n_y$) is odd.
\end{itemize}
\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{../../plots/Boundary_Cond_Sketch.png}
\end{center}
\caption{Sketch for patches of copied (and mirrored) positions for periodic (A) and Neumann (B) boundary conditions.}
\label{Bound_Cond_Patches}
\end{figure}
Therefore, the entries of the operator for periodic boundary conditions $\hat{\psi}_{per.}$ could be calculated by
\begin{equation}
\psi_{ij,per.} = \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} (d(\mathbf{x}_{neur,i},\mathbf{x}_{neur,i} + L \cdot (n_x,n_y))) \; .
\label{Psi_entries_periodic_bound}
\end{equation}
For the Neumann boundary condition, we found
\begin{align}
\psi_{ij,neum.} &= \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} \left( d(\mathbf{x}_{neur,i},M(\mathbf{x}_{neur,i}) + L \cdot (n_x,n_y)) \right) \label{Psi_entries_neumann_bound} \\
M(\mathbf{x}) &\equiv  
 \begin{pmatrix}
  (-1)^{n_x} & 0\\
  0 & (-1)^{n_y}
 \end{pmatrix}
\mathbf{x} + L \cdot (mod(\left| n_x\right|,2),mod(\left| n_y\right|,2)) \; . \label{Mirror_Operator}
\end{align}
Note that the shift was applied to the second position. Theoretically, $n_x$,$n_y$ are to be iterated over all integers and in this case it is irrelevant whether to shift the first or the second position. We obviously had to limit the amount of elements of the sum for our calculations, such that copies that are shifted further away can be neglected. For the sake of performance the number of terms in the sum was adjusted to suffice the given diffusion constant.

@@

\subsubsection{Low-Density Neighborhoods correlate with strong outgoing Connections}
In Section \ref{Section_Mean_outgoing_Weights} we stated that diffusive homeostasis allowed the network do develop few neurons with exceptionally strong outgoing weights. Furthermore, we found this effect to be present in neurons with an highly above-average firing rate. Moreover, the outcomes presented in Section \ref{Section_Rand_Mat_vs_Sim} suggest that the steady state of firing rates is strongly influenced by the spatial structure of excitatory neurons. Combining these two qualities led us the conclusion that one should be able to observe some relation between spatial structure and the emergence of "driver neurons". More specific, regions of neurons with above average firing rates should develop stronger outgoing connections. To get an intuitive understanding of what characterizes regions of above-average firing rates we interpreted the homeostatic condition in the high-neuron-density limit as an approximation of a continuous integral equation:
\begin{equation}
\int d^2 x' \rho(\mathbf{x}') r(\mathbf{x}') \psi(|\mathbf{x}-\mathbf{x}'|) = NO_0 \; , \; \forall \, \mathbf{x}
\label{Density_Limit_Hom}
\end{equation}
where $\mathrm{\rho(\mathbf{x})}$ is the local neuron density and $\mathrm{r(\mathbf{x})}$ the local firing rate. Here we assumed that our tissue is of infinite size and thus the integral is carried out over $\mathbb{R}^2$. In this form the solution for $\mathrm{r(\mathbf{x})}$ is quite trivial:
\begin{equation}
r(\mathbf{x}) = \frac{NO_0}{\rho(\mathbf{x}) \int d^2 x' \psi(|\mathbf{x}'|)} \propto \frac{1}{\rho(\mathbf{x})} \; .
\label{Density_Limit_Hom_Solution}
\end{equation}
Combined with the fact that the logarithm of outgoing excitatory weights correlates with excitatory firing rate (see Figure \ref{Out_Weight_vs_F}), this led us to the conclusion that low density regions should correlate with strong outgoing weights and vice versa. We approximated the local neuron density by means of a a Gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$ to test this hypothesis. Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} shows the results. Despite a certain amount of randomness, our prediction was indeed verified. 
\begin{figure}
%% density_vs_sum_out_weights.py
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights.png}
%% density_vs_sum_out_weights_points.py
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights_points.png}
\caption{(A) and (B): Scatter-plot of inverse of local neuron density (calculated by convolution with a gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$) and decadic logarithm of outgoing excitatory weights. Note the spatial correlation in high/low-density regions. (C) and (D): The decadic logarithm of outgoing weights correlates with inverse neuron density (corr. coeff. of 0.513).}
\label{Inverse_Dens_vs_Sum_Out_Weights}
\end{figure}

@@

\subsection{The Firing Rate Distribution} \label{Fir_Rate_Dist_Discussion_Section}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics we developed analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches of understanding the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behavior of recurrent networks are largely based on the assumption that the excitability of neurons is not individually fine-tuned but rather randomly distributed and adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this canonical approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal many differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behavior.

>>>>

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
%\usepackage{geometry}
%\geometry{
%a4paper,
%left=2.0in,
%right=1.0in}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}

@@

\subsubsection{Mean outgoing Weights}\label{Section_Mean_outgoing_Weights}
Since we only implemented synaptic normalization for the sum of \emph{ingoing} weights, the total or average strength of \emph{outgoing} connections may vary from cell to cell. In a sense, the strength of outgoing connections per neuron can be regarded as a measure of how "influential" a neuron's activity is with respect to other neurons in the network. Effenberger et al. have shown in computational studies that these "driver neurons" form highly active and interconnected subnetworks \cite{Effenberger_2015}, an observation that is backed up by experimental studies \cite{Yassin_Subnetworks_2010,Eckmann_Leader_Neurons_2008}. Figure \ref{Out_Weight_Mean_Quantile} shows that diffusive homeostasis allows for the emergence of a small group of highly influential neurons, which resembles the findings in \cite{Effenberger_2015}. A comparison of statistics generated with shuffled versions of the weight matrices illustrates that above-chance exceptionally strong mean outgoing weights were indeed present. We noted further that the difference between shuffled matrices corresponding to non-diffusive and diffusive homeostasis is in line with our observation from Figure \ref{Weight_Dist}, which indicated a broader distribution of weights for diffusive homeostasis.
\begin{figure}
%%% plot_sum_outgoing_weights_dist.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/outgoing_weights_distribution.png}
\caption{Mean of outgoing weights of excitatory neurons in order sorted by magnitude. Dashed lines are means obtained by randomly shuffling synaptic weights. Data was taken from 25 weight matrices per line, representing weights within $\mathrm{t=1000-1500\,s}$.}
\label{Out_Weight_Mean_Quantile}
\end{figure}

Effenberger et al. identified the presence of inhibitory STDP as a necessary mechanism for the development of this feature, in contrast to our model. On a deeper explanatory level, the authors argue that inhibitory STDP is indirectly involved: It weakens inhibitory currents for some neurons, which cause these neurons to have more above-average firing rates. In turn, synapses between highly active presynaptic cells and postsynaptic neurons with lower activity are known to be subject to long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}. With these causal relations in mind, we can argue that diffusive homeostasis effectively embodies a similar functional role as inhibitory STDP in \cite{Effenberger_2015} by allowing for the presence of excitatory cells with above-average activity. We tested this relationship by plotting the mean outgoing weights against the average firing rate, see Figure \ref{Out_Weight_vs_F}. A strong heterogeneity of firing rates allows for the development of few neurons with comparably strong mean outgoing weights, while the distribution of mean weights resulting from a narrow distribution is limited to a smaller range. We also tested the case of instant diffusion. Though firing rates span across a similar range, mean weights do not reach as high values as in the case of diffusion at a finite rate.
\begin{figure}
%%% out_weight_vs_f.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/mean_out_weights_vs_rate.png}
\caption{Mean of outgoing excitatory weights at $\mathrm{t=1500\,s}$ plotted against mean presynaptic firing rate (averaged over $\mathrm{t=700-1000\,s}$) in regular (A) and logarithmic space (B). Each plot corresponds to a presynaptic excitatory cell.}
\label{Out_Weight_vs_F}
\end{figure}
Furthermore, it is interesting to note that we found a roughly linear relationship on a logarithmic weight scale. Figure \ref{Out_Weight_vs_F} only depicts data from simulations including a distance dependent connection profile, since removing this topology did not result in noticeable differences.
\newpage
\section{Further Analysis of the Firing Rate Distribution}
The results of section \ref{Fir_Dist_Section} were positive on a preliminary level regarding our goal of achieving a broad and skewed firing rate distribution. Following this, we tried to further gain insight into the actual shape of the resulting distribution. 
\subsection{Review of a Dynamic Mean-Field Model}\label{Mean_Field_Review_Section}
Sweeney et al. attempted to predict the shape of the firing rate distribution by means of a simplified mean-field approximation of diffusion \cite{Sweeney_Paper}. The main simplification within homeostatic feedback was the replacement of Equation \ref{Theta_dyn} by
\begin{equation}
\dot{V_{t,i}}(t) = \frac{1}{\tau_{HIP}} \left( (1-\alpha)\frac{\phi_i-\phi_0}{\phi_i} +\alpha \frac{\langle \phi \rangle-\phi_0}{\langle \phi \rangle} \right)\;. 
\label{diff_hom_simpl_sweeney}
\end{equation}
$\mathrm{\alpha \: \epsilon \: [0,1]}$ thereby acted as a parameter that determines the mixture between single-neuron homeostasis ($\mathrm{\alpha=0}$) and the limit of quasi instantaneous spreading of the diffusive signal across the population ($\mathrm{\alpha=1}$). $\mathrm{\phi_i}$ and $\mathrm{\mathrm{\langle \phi \rangle}}$ represent individual firing rates and their overall mean.

@@

Generally speaking, no matter how the actual shape of the numeric solution in the equilibrium at a constant production rate looks like, it must be of the form
\begin{equation}
NO_i(\mathbf{x}) = r_i \cdot \psi (|\mathbf{x}_{neur,i}-\mathbf{x}|) \; . \label{general_diff_interaction}
\end{equation}
The full solution is then
\begin{equation}
NO(\mathbf{x}) = \sum_i NO_i(\mathbf{x})\;.
\label{full_sol_diff_equil}
\end{equation}
By defining
\begin{equation}
\psi_{ij} \equiv \psi_{ij} \equiv \psi (|\mathbf{x}_{neur,i}-\mathbf{x}_{neur,i}|)
\label{interact_matrix_elements}
\end{equation}
we could express the condition \eqref{NO_equil_cond} as
\begin{equation}
\sum_j \psi_{ij}\cdot r_j = NO_0
\label{NO_equil_cond_interact_matrix}
\end{equation}
or, as an operator
\begin{align}
\hat{\psi}\mathbf{r} &= NO_0 \mathbf{n} \label{NO_equil_cond_interact_matrix_operator} \\
\mathbf{n}&\equiv (1,1,...,1) \; .
\end{align}
The problem of finding the steady-state solution of the homeostatic constraint thus reduced to inverting $\hat{\psi}$:
\begin{equation}
\mathbf{r} = NO_0 \hat{\psi}^{-1} \mathbf{n}
\label{NO_euqil_cond_interact_matrix_operator_solve}
\end{equation}
Still, we had to find a modified, non-diverging version of $\mathrm{\psi (|\mathbf{x}_{neur,i}-\mathbf{x}|)}$ to acquire any prediction from this model. It had to retain the shape given by \eqref{solution_diff_equil_bessel} for larger distances but approach the correct numeric error value at the origin, determined by the spacing of the numeric grid. We solved this problem by the following expression:
\begin{equation}
\psi_{approx.} \equiv \frac{1}{\left(\frac{1}{\psi_0^\varepsilon} + \frac{1}{\psi_{point}^\varepsilon}\right)^{\frac{1}{\varepsilon}}}
\label{Numeric_Solution_Expression_Trick}
\end{equation}
where $\mathrm{\varepsilon}$ determines the "smoothness" of transition between $\mathrm{\psi}$ and the cutoff value $\mathrm{\psi_0}$. We chose $\mathrm{\varepsilon=10}$ for all further calculations. We took the simple approach of interpreting this value as a mean of the analytic solution across the area covered by the corresponding grid cell to find an expression for $\mathrm{\psi_0}$. As an additional simplification, we substituted the necessary integration over the square grid cell by a circular area of equal size around the source. This calculation yielded
\begin{equation}
\psi_0 = \gamma \frac{1-h\sqrt{\frac{\lambda}{\pi D}} K_1\left( h\sqrt{\frac{\lambda}{\pi D}}\right) }{h^2 \lambda}
\label{Numeric_Grid_Bessel_Approx}
\end{equation}
where $\mathrm{h}$ is the spatial resolution of the grid cells. Figure \ref{Psi_Approximation_Compare} shows a comparison between the numerically calculated solution and the expression given by \ref{Numeric_Solution_Expression_Trick}. The approximation fitted very well for nonzero values, as expected. The value at the critical point in the origin was slightly underestimated by \eqref{Numeric_Grid_Bessel_Approx}, but nonetheless fitted well into the overall shape.
\begin{figure}
%% compare_num_diff_steady_vs_analytic.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/Psi_Interact_Approximation.png}
\end{center}
\caption{Numerical steady-state solution of Equation \eqref{simple_NO_dyn_with_diff_equil_helmholtz} (blue, $\mathrm{h=10\;\mu m}$) and its approximation (dashed) given by \eqref{Numeric_Solution_Expression_Trick}. The numerical simulation was carried out on a 2d-grid and the curve represents a cut through the origin in x-direction.}
\label{Psi_Approximation_Compare}
\end{figure}
We took this approximation as a basis for further calculations of the interaction-matrix $\mathrm{\hat{\psi}}$.

By simply calculating all matrix elements of $\mathrm{\hat{\psi}}$ by means of Equation \eqref{interact_matrix_elements}, one would neglect the finite boundaries of the system, which would cause neurons close to the edge to "bleed" into empty space. This in turn would cause the solution of \eqref{NO_equil_cond_interact_matrix_operator} to contain an over-representation of high firing rates since close-to-the-edge neurons would need to compensate for their lack of neighbors. Thus, we had to account for the boundary conditions used in the network simulation. We simulated the network mostly with Neumann boundary conditions as well as periodic boundaries, as described in section \ref{methods}. Both types can be modeled by extending the neurons' population through spatially shifted and mirrored versions of the base population (see Figure \ref{Bound_Cond_Patches}):
\begin{itemize}
\item Periodic boundary conditions are induced by copies of the neurons' positions shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$.
\item Zero flux through the boundaries can be achieved by copied positions being shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$ \textit{and} mirrored in $x$($y$)-direction if $n_x$($n_y$) is odd.
\end{itemize}
\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{../../plots/Boundary_Cond_Sketch.png}
\end{center}
\caption{Sketch for patches of copied (and mirrored) positions for periodic (A) and Neumann (B) boundary conditions.}
\label{Bound_Cond_Patches}
\end{figure}
Therefore, the entries of the operator for periodic boundary conditions $\hat{\psi}_{per.}$ could be calculated by
\begin{equation}
\psi_{ij,per.} = \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} (|\mathbf{x}_{neur,i}-\mathbf{x}_{neur,i} - L \cdot (n_x,n_y)|) \; .
\label{Psi_entries_periodic_bound}
\end{equation}
For the Neumann boundary condition, we found
\begin{align}
\psi_{ij,neum.} &= \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} \left(|\mathbf{x}_{neur,i}-M(\mathbf{x}_{neur,i}) - L \cdot (n_x,n_y)| \right) \label{Psi_entries_neumann_bound} \\
M(\mathbf{x}) &\equiv  
 \begin{pmatrix}
  (-1)^{n_x} & 0\\
  0 & (-1)^{n_y}
 \end{pmatrix}
\mathbf{x} + L \cdot (mod(\left| n_x\right|,2),mod(\left| n_y\right|,2)) \; . \label{Mirror_Operator}
\end{align}
Note that the shift was applied to the second position. Theoretically, $n_x$,$n_y$ are to be iterated over all integers and in this case it is irrelevant whether to shift the first or the second position. We obviously had to limit the amount of elements of the sum for our calculations, such that copies that are shifted further away can be neglected. For the sake of performance the number of terms in the sum was adjusted to suffice the given diffusion constant.

@@

\subsubsection{Low-Density Neighborhoods correlate with strong outgoing Connections}
In Section \ref{Section_Mean_outgoing_Weights} we stated that diffusive homeostasis allowed the network do develop few neurons with exceptionally strong outgoing weights. We found this effect to be present in neurons with an highly above-average firing rate. Moreover, the outcomes presented in Section \ref{Section_Rand_Mat_vs_Sim} suggest that the steady state of firing rates is strongly influenced by the spatial structure of excitatory neurons. Combining these two qualities led us the conclusion that one should be able to observe some relation between spatial structure and the emergence of "driver neurons". More specific, regions of neurons with above average firing rates should develop stronger outgoing connections. To get an intuitive understanding of what characterizes regions of above-average firing rates we interpreted the homeostatic condition in the high-neuron-density limit as an approximation of a continuous integral equation:
\begin{equation}
\int d^2 x' \rho(\mathbf{x}') r(\mathbf{x}') \psi(|\mathbf{x}-\mathbf{x}'|) = NO_0 \; , \; \forall \, \mathbf{x}
\label{Density_Limit_Hom}
\end{equation}
where $\mathrm{\rho(\mathbf{x})}$ is the local neuron density and $\mathrm{r(\mathbf{x})}$ the local firing rate. Here we assumed that our tissue is of infinite size and thus the integral is carried out over $\mathbb{R}^2$. In this form the solution for $\mathrm{r(\mathbf{x})}$ is quite trivial:
\begin{equation}
r(\mathbf{x}) = \frac{NO_0}{\rho(\mathbf{x}) \int d^2 x' \psi(|\mathbf{x}'|)} \propto \frac{1}{\rho(\mathbf{x})} \; .
\label{Density_Limit_Hom_Solution}
\end{equation}
Combined with the fact that the logarithm of outgoing excitatory weights correlates with excitatory firing rate (see Figure \ref{Out_Weight_vs_F}), this led us to the conclusion that low density regions should correlate with strong outgoing weights and vice versa. We approximated the local neuron density by means of a a Gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$ to test this hypothesis. Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} shows the results. Despite a certain amount of randomness, our prediction was indeed verified. 
\begin{figure}
%% density_vs_sum_out_weights.py
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights.png}
%% density_vs_sum_out_weights_points.py
\includegraphics[width=\textwidth]{../../plots/scatter_density_out_weights_points.png}
\caption{(A) and (B): Scatter-plot of inverse of local neuron density (calculated by convolution with a gaussian kernel with $\mathrm{\sigma = 50\, \mu m}$) and decadic logarithm of outgoing excitatory weights. Note the spatial correlation in high/low-density regions. (C) and (D): The decadic logarithm of outgoing weights correlates with inverse neuron density (corr. coeff. of 0.513).}
\label{Inverse_Dens_vs_Sum_Out_Weights}
\end{figure}

@@

\subsection{The Firing Rate Distribution} \label{Fir_Rate_Dist_Discussion_Section}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics we developed an analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches of understanding the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behavior of recurrent networks are largely based on the assumption that the excitability of neurons is not individually fine-tuned but rather randomly distributed and adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this canonical approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal many differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behavior.

Sweeney et al. claimed that their implementation of diffusive homeostasis enables individual neurons to vary their firing rates more flexible compared to single-neuron homeostatic models, especially allowing for sustained encoding of different levels of external input. Interestingly, we could not confirm this, at least within the tests that we performed and the given set of standard parameters. Overall, the resulting behavior of the excitatory population rather appeared to us as a form of single-neuron homeostasis whose individual targets are predetermined by the quenched random spatial structure of the respective cells. In particular, our result depicted in Figure \ref{Ext_Input_Switch} did not support the notion of separability of inputs. Of course, this impression only applies to the case where the aforementioned mixture of influence between spatial and synaptic topology is dominated by the former. By analyzing network activity based on our analytic model we found that indeed only a single homeostatic fixed point exists for finite diffusion constants. On a practical level though, the gradual decorrelation with this theoretical state of activity upon larger values of $\mathrm{D}$ was presumably caused by synaptic turnovers which change the synaptic input. Fluctuations in activity were more slowly and less accurately compensated in the case of larger diffusion constants. However, for the standard choice of parameters (being roughly based on experimental data) the resulting activity was indeed strongly bound to the spatial structure of excitatory cells. Naturally, this raises the question of biological plausibility. To our knowledge, no experimental study exists that relates measurements of spontaneous activity to local fluctuations of neuronal densities. Ivenshitz and Segal studied the influence of (mean) neuronal density onto neuronal activity in a culture of hippocampal neurons \cite{Ivenshitz_2010}. Lower densities resulted in a more irregular activity, with stronger but less frequent bursting events. However, it is hard to draw any conclusions from this result since our network was tuned to exhibit non-bursting activity.
